### 2023-07-20 chapters/paper_hierarchical.tex
Some formatting

\chapter[hierarchical vaes know what they don't know]{Hierarchical VAEs Know What They Don't Know}
\label{chp:hierarchical}

@@

\section{Introduction}
\begin{figure}[t]
    \centering
    \includegraphics[width=0.7\columnwidth]{paper_hierarchical/reconstructions-front-page-4-samples-2-recons2.pdf}
    \caption[Reconstructions of a hierarchical VAE trained on FashionMNIST.]{%
        Reconstructions using a hierarchical VAE trained on FashionMNIST.
        Reconstruction quality of OOD data is comparable to in-distribution data, resulting in high likelihoods and poor OOD discrimination.
        By sampling the $k$ bottom-most latent variables from the conditional prior distribution $p(\zb_{\geq l}|\zb_{>l})$ (latent reconstructions) instead of the approximate posterior $q(\zb_{>l}|\zb_{<l})$, the model reconstructs from the training distribution resulting in lower $p(\xb|\zb)$ for OOD data.
    }
    \label{fig:reconstructions-fashionmnist}
\end{figure}
% First paragraph: Explains why OOD detection is important in a broad perspective
The reliability and safety of machine learning systems applied in the real-world is contingent on the ability to detect when an input is different from the training distribution. %, an anomaly.
Supervised classifiers built as deep neural networks are well-known to misclassify such \textit{out-of-distribution} (OOD) inputs to known classes with high confidence \cite{goodfellow_explaining_2015, nguyen_deep_2015}.
Several approaches have been suggested to equip deep classifiers with OOD detection capabilities \cite{hendrycks_baseline_2017, lakshminarayanan_simple_2017, hendrycks_deep_2019, devries_learning_2018}.
% However, such methods are inherently supervised and require in-distribution labels or examples of OOD data from an a priori known distribution which limit their applicability.
But, such methods are inherently supervised and require in-distribution labels or examples of OOD data limiting their applicability and generality.

@@

\section{Why does OOD detection fail?}\label{sec:why-does-ood-fail}
\begin{figure}[t]
    \centering
    \includegraphics[width=1\columnwidth]{paper_hierarchical/feature-correlation-heatmap2.pdf}
    \vspace{0mm}
    \caption[Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.]{
    Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.
    We compute the correlation between the representations of the two different models given the same data, FashionMNIST (top) and MNIST (bottom).
    %We reduce the representations to a single number by summing over all dimensions and compute the correlation between these numbers for 256 examples.
    %To reduce de-correlation due to the stochastic latent variables, we encode using the mode of the approximate posterior distributions.
    }
    \vspace{0mm}
    \label{fig:correlations-heatmap}
\end{figure}

@@

Deep learning models are generally believed to form hierarchies of representations that range from low-level features to more conceptual ones related to semantics \cite{bengio_representation_2013}.
This has also been observed within deep generative models \cite{maaloe_biva_2019, child_very_2021}.
%Recent research \cite{maaloe_biva_2019, child_very_2021, vahdat_nvae_2020}, has shown that deep latent variable models are able to learn rich hierarchical representations with increasing levels of abstraction and global structure towards the top of the hierarchy.
For image data there is a trend that the low-level features are quite similar across models (edge detectors, etc.). This raises the question to what extend such features are relevant when detecting OOD data, also suggested  by \cite{nalisnick_deep_2019} and examined for Glow and PixelCNN in \cite{schirrmeister_understanding_2020}.
To investigate, we train two hierarchical VAEs (\cref{sec:background-hie-VAE}) on FashionMNIST and MNIST, respectively, and compute the between-models correlation of the extracted features of in-distribution data and OOD data.
The result appears in \cref{fig:correlations-heatmap}.
We observe that features extracted in the early layers (low-level features) correlate strongly between the two models, and that this correlation drops as we get into later layers.
This suggests that low-level features do not carry much information for OOD detection.

To shed further light on the impact of semantic versus low-level features, we look at model reconstructions of images with a hierarchical VAE (\cref{fig:biva-reconstructions-celeba}).
To study the feature hierarchy, we replace the inference distribution with the corresponding conditional prior in the first layers of the model to see what information is lost.
We observe that as more layers rely on the prior, more details are lost.
Sunglasses, which are uncommon, are first replaced by more common glasses, and then finally disappear.
%Likewise, a long-haired man eventually becomes woman.
This suggests that as we fall back to the conditional priors of each layer, we are pushed closer to local modes of the modeled distribution.

\begin{figure}[t]
    \centering
    %\vspace{0.17cm}
    \includegraphics[width=1\columnwidth]{paper_hierarchical/biva_reconstructions.pdf}
    % https://docs.google.com/drawings/d/1UXdA3c18Oek9_env1vEB9mqOAHZ-tu0ijWmvqmapWCk/edit?usp=sharing
    \vspace{0mm}
    \caption[Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables.]{Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables  \cite{maaloe_biva_2019}.
    The higher the latent variable, the more the reconstructions fall into the mode of the learned distribution.
    It is more common to wear regular glasses than sunglasses but most common not to wear glasses at all.
    A man with long hair collapses into the mode of the more common long-haired woman.}
    \vspace{0mm}
    \label{fig:biva-reconstructions-celeba}
\end{figure}
Finally, we look at reconstructions of out-of-distribution data.
\cref{fig:reconstructions-fashionmnist} illustrates that MNIST data is surprisingly well reconstructed by a hierarchical VAE trained on FashionMNIST.
Similar results have been found elsewhere \cite{xiao_likelihood_2020}.
We repeat the previous experiment and replace inference distributions by their corresponding conditional prior, and now observe that reconstructions from higher latent layers become increasingly similar to the data on which the model was trained.
The reliance on conditional priors seems to prevent accurate reconstruction of out-of-distribution data.
Some details are lost on in-distribution data too, but the distinction between that and out-of-distribution data becomes more clear.

@@

\subsection{Variational autoencoders}
The variational autoencoder (VAE) \cite{kingma_autoencoding_2014, rezende_stochastic_2014} is a framework for constructing deep generative models defined by an observed variable $\mathbf{x}$ and a stochastic latent variable $\mathbf{z}$.
Typically, a neural network with parameters $\thetab$ is chosen to parameterize the generative distribution $p_\thetab(\xb,\zb)=p_\thetab(\xb|\zb)p(\zb)$, where the prior $p(\zb)$ is commonly a standard Gaussian $\mathcal{N}(\0, \Ib)$.
The true posterior $p(\zb|\xb)$ is generally not analytically tractable and is approximated by a variational distribution $q_\phib(\zb|\xb)$ parameterized via another neural network with parameters $\phib$. The approximate posterior $q_\phib(\zb|\xb)$ is most often  a diagonal covariance Gaussian.
The model parameters $\thetab$ and variational parameters $\phib$ are jointly optimized by maximizing the \textit{evidence lower bound} (ELBO),
\begin{equation}\label{eq:elbo}
    \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}(\xb; \thetab, \phib)\ .
\end{equation}
For brevity, we will denote $\mathcal{L}(\xb; \thetab, \phib)$ as $\mathcal{L}(\xb)$ or $\mathcal{L}$. The reparameterization trick is used to backpropagate gradients through the stochastic latent variables with low variance.

@@

\subsection{Hierarchical variational autoencoders}\label{sec:background-hie-VAE}
Hierarchical VAEs are a family of probabilistic latent variable models which extends the basic VAE by introducing a hierarchy of $L$ latent variables $\zb=\zb_1, \dots, \zb_L$.
The most common generative model is defined from the top down as $p_\thetab(\xb|\zb)=p(\xb|\zb_1)p_\thetab(\zb_1|\zb_2)\cdots p_\thetab(\zb_{L-1}|\zb_L)$.
The inference model can then be defined in two ways respectively referred to as \textit{bottom-up} \cite{burda_importance_2016}
\begin{equation}
    %q_\phib(\zb|\xb)=q_\phib(\zb_1|\xb)q_\phib(\zb_2|\zb_1)\cdots q_\phib(\zb_L|\zb_{L-1})
    q_\phib(\zb|\xb) = q_\phib(\zb_1|\xb)\textstyle\prod_{i=2}^{L} q_\phib(\zb_i|\zb_{i-1})
\end{equation}
and \textit{top-down} \cite{sonderby_ladder_2016}
\begin{equation}
    % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb_{L-1}|\zb_L)\cdots q_\phib(\zb_2|\zb_1),
    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
\end{equation}
Regardless of the choice of inference model, a hierarchical VAE is still trained using the ELBO \eqref{eq:elbo}.

@@

\subsection{Out-of-distribution detection}\label{sec:background-ood-detection}
% Intro to alternative scores
So far, no reliable direct likelihood-based method has been found for fully unsupervised deep generative model OOD detection.
A major line of work considers developing new scores that are more reliable than the likelihood.
This includes the \textit{typicality} test presented by \textcite{nalisnick_detecting_2019} which is an OOD detection test based on the typicality of a batch of potentially OOD examples.
This approach however requires a batch of examples from the same class (OOD or not) which limits its practical applicability.
In \textcite{ren_likelihood_2019}, the \textit{likelihood ratio} between a primary model and a background model was shown to be an effective score for OOD detection.
However, to train the background model, the in-distribution data is perturbed via a data augmentation technique that is designed with knowledge about the confounding factors between the in-distribution data and the OOD data. Furthermore, it is tuned towards high performance on a known OOD dataset.
\textcite{serra_input_2020} take a similar approach and attribute the failure to detect OOD data to the high influence of the input complexity on the likelihood and choose a generic lossless compression algorithm as the background model.
Although this method gives good results, no single best choice of compression algorithm exists for all types of OOD data, and any particular choice encodes prior knowledge about the data into the detection method.
Both these methods can be seen as correcting for low-level features of the OOD data being assigned high model likelihood by using a second model focused exclusively on these features.

@@

\section{OOD detection with hierarchical VAEs}
\subsection{A bound for semantic OOD detection}
If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \cref{fig:reconstructions-fashionmnist}-\ref{fig:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ from \eqref{eq:elbo} relies on the KL-term for OOD detection. For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ can be expressed by a hierarchical sum% over the hierarchy
% \begin{multline}
%     \mathcal{L}(\xb) =  \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \log p_\thetab(\xb|\zb_1) \\
%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} 
%                      + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
% \end{multline}
% \begin{multline}
%     D_{\mathrm{KL}}( q(\zb|\xb) || p(\zb)) = \\
%     \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
% \end{multline}
\begin{equation}
    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] \ .
\end{equation}
In general, the absolute log-ratios grow with $\mathrm{dim}(\zb_i)$ as the individual log probability terms are computed by summing over the dimensionality of $\zb_i$.
This means that the value of the KL-term is dominated by terms where $\zb_i$ is high-dimensional. We refer to Appendix C for a more detailed argument.
Since hierarchical VAEs are generally constructed with a bottleneck type structure, the terms corresponding to latent variables towards the top of the hierarchy will have a vanishing influence on the value of the KL-term.
However, as the semantic information most relevant for OOD detection has a tendency to be represented in the top-most latent variables, this makes OOD detection using the regular ELBO difficult, even for state-of-the-art models.
This behavior has also been reported by \textcite{xiao_likelihood_2020}.

To shift the ELBO from primarily being based on the approximate posterior of the lowest latent variables to instead focus on the conditional prior, \textcite{maaloe_biva_2019} introduced slightly different likelihood lower bound defined as
\begin{equation}\label{eq:biva->k}
    \mathcal{L}^{>k} = \mathbb{E}_{p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb_{>k})}{q_\phib(\zb_{>k}|\xb)} \right]
\end{equation}
where $k\in\{0,1,\dots,L\}$ (see Appendix for the derivation).
We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
Importantly, this has the effect that the data likelihood $p(\xb|\zb)$ is dependent on the approximate posterior through a latent variable $\zb_{k+1}$ different from $\zb_1$ for all $k \geq 1$.
Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\zb_k$ of the hierarchical VAE.
Hence, we can now test how well the input $\xb$ is reconstructed from each latent variable.
The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\zb_{>k}$, the bound is the regular ELBO while for the latent variables $\zb_{\leq k}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.

\subsection{A likelihood-ratio score for all feature levels}
While the $\mathcal{L}^{>k}$ bound provides a score for performing semantic OOD detection, it still relies on the data space likelihood function (see equation \eqref{eq:likelihoods-as-exact} below), which is known to be problematic for OOD detection (\cref{sec:background-ood-detection}). To alleviate this, we phrase OOD detection as a likelihood ratio test of being \emph{semantically} in-distribution.
A standard likelihood ratio test \cite{buse_likelihood_1982} suggests to consider the ratio between the associated likelihoods, which we can approximate on a log-scale by the corresponding lower bounds $\mathcal{L}$ and $\mathcal{L}^{>k}$,
\begin{equation}\label{eq:llr-as-difference-in-likelihoods}
    % LLR = - 2 log(L0 / L1)
    % where L0 < L1 so log(L0 / L1) < 0 and LLR > 0
    % LLR = 2 log(L1 / L0)
    %     = 2 (log(L1) - log(L0))
    %     = 2 (ELBO - L^{>k})  % LLR^{>a,>b} where a=0
    % L1 = ELBO
    % L0 = L^{>k}
    LLR^{>k}(\xb) = \mathcal{L}(\xb) - \mathcal{L}^{>k}(\xb) \ .
\end{equation}
%This likelihood ratio contrasts the ELBO with $\mathcal{L}^{>k}$ for a variable choice of $k$.
Since, empirically, $\mathcal{L}\geq\mathcal{L}^{>k}$, the ratio is always positive as is standard for likelihood ratio tests.
A low value of $LLR^{>k}(\xb)$ means that the ELBO and $\mathcal{L}^{>k}$ are almost equally tight for the data.
On the contrary, a high value indicates that $\mathcal{L}^{>k}$ is looser on the data than the ELBO; hence, the data may be OOD.

>>>>

\chapter[hierarchical vaes know what they don't know]{Hierarchical VAEs Know What They Don't Know}
\label{chp:paper-hierarchical}
% \ifthenelse{\equal{\skippaperbody}{true}}{}{

@@

\section{Introduction}
\begin{figure}[t]
    \centering
    \includegraphics[width=0.7\columnwidth]{paper_hierarchical/reconstructions-front-page-4-samples-2-recons2.pdf}
    \caption[Reconstructions of a hierarchical VAE trained on FashionMNIST.]{%
        Reconstructions using a hierarchical VAE trained on FashionMNIST.
        Reconstruction quality of OOD data is comparable to in-distribution data, resulting in high likelihoods and poor OOD discrimination.
        By sampling the $k$ bottom-most latent variables from the conditional prior distribution $p(\zb_{\geq l}|\zb_{>l})$ (latent reconstructions) instead of the approximate posterior $q(\zb_{>l}|\zb_{<l})$, the model reconstructs from the training distribution resulting in lower $p(\xb|\zb)$ for OOD data.
    }
    \label{fig_hierarchical:reconstructions-fashionmnist}
\end{figure}
% First paragraph: Explains why OOD detection is important in a broad perspective
The reliability and safety of machine learning systems applied in the real-world is contingent on the ability to detect when an input is different from the training distribution. %, an anomaly.
Supervised classifiers built as deep neural networks are well-known to misclassify such \textit{out-of-distribution} (OOD) inputs to known classes with high confidence \cite{goodfellow_explaining_2015, nguyen_deep_2015}.
Several approaches have been suggested to equip deep classifiers with OOD detection capabilities \cite{hendrycks_baseline_2017, lakshminarayanan_simple_2017, hendrycks_deep_2019, devries_learning_2018}.
% However, such methods are inherently supervised and require in-distribution labels or examples of OOD data from an a priori known distribution which limit their applicability.
But, such methods are inherently supervised and require in-distribution labels or examples of OOD data limiting their applicability and generality.

@@

\section{Why does OOD detection fail?}\label{sec_paper_hierarchical:why-does-ood-fail}
\begin{figure}[t]
    \centering
    \includegraphics[width=1\columnwidth]{paper_hierarchical/feature-correlation-heatmap2.pdf}
    \vspace{0mm}
    \caption[Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.]{
    Absolute correlations between data representations in all layers of the inference network of a hierarchical VAE trained on FashionMNIST and of another trained on MNIST.
    We compute the correlation between the representations of the two different models given the same data, FashionMNIST (top) and MNIST (bottom).
    %We reduce the representations to a single number by summing over all dimensions and compute the correlation between these numbers for 256 examples.
    %To reduce de-correlation due to the stochastic latent variables, we encode using the mode of the approximate posterior distributions.
    }
    \vspace{0mm}
    \label{fig_hierarchical:correlations-heatmap}
\end{figure}

@@

Deep learning models are generally believed to form hierarchies of representations that range from low-level features to more conceptual ones related to semantics \cite{bengio_representation_2013}.
This has also been observed within deep generative models \cite{maaloe_biva_2019, child_very_2021}.
%Recent research \cite{maaloe_biva_2019, child_very_2021, vahdat_nvae_2020}, has shown that deep latent variable models are able to learn rich hierarchical representations with increasing levels of abstraction and global structure towards the top of the hierarchy.
For image data there is a trend that the low-level features are quite similar across models (edge detectors, etc.). This raises the question to what extend such features are relevant when detecting OOD data, also suggested  by \cite{nalisnick_deep_2019} and examined for Glow and PixelCNN in \cite{schirrmeister_understanding_2020}.
To investigate, we train two hierarchical VAEs (\cref{sec_paper_hierarchical:background-hie-VAE}) on FashionMNIST and MNIST, respectively, and compute the between-models correlation of the extracted features of in-distribution data and OOD data.
The result appears in \cref{fig_hierarchical:correlations-heatmap}.
We observe that features extracted in the early layers (low-level features) correlate strongly between the two models, and that this correlation drops as we get into later layers.
This suggests that low-level features do not carry much information for OOD detection.

To shed further light on the impact of semantic versus low-level features, we look at model reconstructions of images with a hierarchical VAE (\cref{fig_hierarchical:biva-reconstructions-celeba}).
To study the feature hierarchy, we replace the inference distribution with the corresponding conditional prior in the first layers of the model to see what information is lost.
We observe that as more layers rely on the prior, more details are lost.
Sunglasses, which are uncommon, are first replaced by more common glasses, and then finally disappear.
%Likewise, a long-haired man eventually becomes woman.
This suggests that as we fall back to the conditional priors of each layer, we are pushed closer to local modes of the modeled distribution.

\begin{figure}[t]
    \centering
    %\vspace{0.17cm}
    \includegraphics[width=1\columnwidth]{paper_hierarchical/biva_reconstructions.pdf}
    % https://docs.google.com/drawings/d/1UXdA3c18Oek9_env1vEB9mqOAHZ-tu0ijWmvqmapWCk/edit?usp=sharing
    \vspace{0mm}
    \caption[Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables.]{Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables  \cite{maaloe_biva_2019}.
    The higher the latent variable, the more the reconstructions fall into the mode of the learned distribution.
    It is more common to wear regular glasses than sunglasses but most common not to wear glasses at all.
    A man with long hair collapses into the mode of the more common long-haired woman.}
    \vspace{0mm}
    \label{fig_hierarchical:biva-reconstructions-celeba}
\end{figure}
Finally, we look at reconstructions of out-of-distribution data.
\cref{fig_hierarchical:reconstructions-fashionmnist} illustrates that MNIST data is surprisingly well reconstructed by a hierarchical VAE trained on FashionMNIST.
Similar results have been found elsewhere \cite{xiao_likelihood_2020}.
We repeat the previous experiment and replace inference distributions by their corresponding conditional prior, and now observe that reconstructions from higher latent layers become increasingly similar to the data on which the model was trained.
The reliance on conditional priors seems to prevent accurate reconstruction of out-of-distribution data.
Some details are lost on in-distribution data too, but the distinction between that and out-of-distribution data becomes more clear.

@@

\subsection{Variational autoencoders}
The variational autoencoder (VAE) \cite{kingma_autoencoding_2014, rezende_stochastic_2014} is a framework for constructing deep generative models defined by an observed variable $\mathbf{x}$ and a stochastic latent variable $\mathbf{z}$.
Typically, a neural network with parameters $\thetab$ is chosen to parameterize the generative distribution $p_\thetab(\xb,\zb)=p_\thetab(\xb|\zb)p(\zb)$, where the prior $p(\zb)$ is commonly a standard Gaussian $\mathcal{N}(\0, \Ib)$.
The true posterior $p(\zb|\xb)$ is generally not analytically tractable and is approximated by a variational distribution $q_\phib(\zb|\xb)$ parameterized via another neural network with parameters $\phib$. The approximate posterior $q_\phib(\zb|\xb)$ is most often  a diagonal covariance Gaussian.
The model parameters $\thetab$ and variational parameters $\phib$ are jointly optimized by maximizing the \textit{evidence lower bound} (ELBO),
\begin{equation}\label{eq_hierarchical:elbo}
    \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}(\xb; \thetab, \phib)\ .
\end{equation}
For brevity, we will denote $\mathcal{L}(\xb; \thetab, \phib)$ as $\mathcal{L}(\xb)$ or $\mathcal{L}$. The reparameterization trick is used to backpropagate gradients through the stochastic latent variables with low variance.

@@

\subsection{Hierarchical variational autoencoders}\label{sec_paper_hierarchical:background-hie-VAE}
Hierarchical VAEs are a family of probabilistic latent variable models which extends the basic VAE by introducing a hierarchy of $L$ latent variables $\zb=\zb_1, \dots, \zb_L$.
The most common generative model is defined from the top down as $p_\thetab(\xb|\zb)=p(\xb|\zb_1)p_\thetab(\zb_1|\zb_2)\cdots p_\thetab(\zb_{L-1}|\zb_L)$.
The inference model can then be defined in two ways respectively referred to as \textit{bottom-up} \cite{burda_importance_2016}
\begin{equation}
    %q_\phib(\zb|\xb)=q_\phib(\zb_1|\xb)q_\phib(\zb_2|\zb_1)\cdots q_\phib(\zb_L|\zb_{L-1})
    q_\phib(\zb|\xb) = q_\phib(\zb_1|\xb)\textstyle\prod_{i=2}^{L} q_\phib(\zb_i|\zb_{i-1})
\end{equation}
and \textit{top-down} \cite{sonderby_ladder_2016}
\begin{equation}
    % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb_{L-1}|\zb_L)\cdots q_\phib(\zb_2|\zb_1),
    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
\end{equation}
Regardless of the choice of inference model, a hierarchical VAE is still trained using the ELBO \cref{eq_hierarchical:elbo}.

@@

\subsection{Out-of-distribution detection}\label{sec_paper_hierarchical:background-ood-detection}
% Intro to alternative scores
So far, no reliable direct likelihood-based method has been found for fully unsupervised deep generative model OOD detection.
A major line of work considers developing new scores that are more reliable than the likelihood.
This includes the \textit{typicality} test presented by \textcite{nalisnick_detecting_2019} which is an OOD detection test based on the typicality of a batch of potentially OOD examples.
This approach however requires a batch of examples from the same class (OOD or not) which limits its practical applicability.
In \textcite{ren_likelihood_2019}, the \textit{likelihood ratio} between a primary model and a background model was shown to be an effective score for OOD detection.
However, to train the background model, the in-distribution data is perturbed via a data augmentation technique that is designed with knowledge about the confounding factors between the in-distribution data and the OOD data. Furthermore, it is tuned towards high performance on a known OOD dataset.
\textcite{serra_input_2020} take a similar approach and attribute the failure to detect OOD data to the high influence of the input complexity on the likelihood and choose a generic lossless compression algorithm as the background model.
Although this method gives good results, no single best choice of compression algorithm exists for all types of OOD data, and any particular choice encodes prior knowledge about the data into the detection method.
Both these methods can be seen as correcting for low-level features of the OOD data being assigned high model likelihood by using a second model focused exclusively on these features.

@@

\section{OOD detection with hierarchical VAEs}
\subsection{A bound for semantic OOD detection}
If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \cref{fig_hierarchical:reconstructions-fashionmnist}-\ref{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb_1)$ will be high for both in- and out-of-distribution data.
Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb_1)] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection. For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) || p(\zb))$ can be expressed by a hierarchical sum% over the hierarchy
% \begin{multline}
%     \mathcal{L}(\xb) =  \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \log p_\thetab(\xb|\zb_1) \\
%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} 
%                      + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
% \end{multline}
% \begin{multline}
%     D_{\mathrm{KL}}( q(\zb|\xb) || p(\zb)) = \\
%     \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] .
% \end{multline}
\begin{equation}
    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb_i|\zb_{i+1})}{q_\phib(\zb_i|\zb_{i-1})} + \log \frac{p_\thetab(\zb_L)}{q_\phib(\zb_L|\zb_{L-1})} \Big] \ .
\end{equation}
In general, the absolute log-ratios grow with $\mathrm{dim}(\zb_i)$ as the individual log probability terms are computed by summing over the dimensionality of $\zb_i$.
This means that the value of the KL-term is dominated by terms where $\zb_i$ is high-dimensional. We refer to Appendix C for a more detailed argument.
Since hierarchical VAEs are generally constructed with a bottleneck type structure, the terms corresponding to latent variables towards the top of the hierarchy will have a vanishing influence on the value of the KL-term.
However, as the semantic information most relevant for OOD detection has a tendency to be represented in the top-most latent variables, this makes OOD detection using the regular ELBO difficult, even for state-of-the-art models.
This behavior has also been reported by \textcite{xiao_likelihood_2020}.

To shift the ELBO from primarily being based on the approximate posterior of the lowest latent variables to instead focus on the conditional prior, \textcite{maaloe_biva_2019} introduced slightly different likelihood lower bound defined as
\begin{equation}\label{eq_hierarchical:biva->k}
    \mathcal{L}^{>k} = \mathbb{E}_{p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb_{>k})}{q_\phib(\zb_{>k}|\xb)} \right]
\end{equation}
where $k\in\{0,1,\dots,L\}$ (see Appendix for the derivation).
We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq_hierarchical:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
Importantly, this has the effect that the data likelihood $p(\xb|\zb)$ is dependent on the approximate posterior through a latent variable $\zb_{k+1}$ different from $\zb_1$ for all $k \geq 1$.
Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\zb_k$ of the hierarchical VAE.
Hence, we can now test how well the input $\xb$ is reconstructed from each latent variable.
The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\zb_{>k}$, the bound is the regular ELBO while for the latent variables $\zb_{\leq k}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.

\subsection{A likelihood-ratio score for all feature levels}
While the $\mathcal{L}^{>k}$ bound provides a score for performing semantic OOD detection, it still relies on the data space likelihood function (see equation \cref{eq_hierarchical:likelihoods-as-exact} below), which is known to be problematic for OOD detection (\cref{sec_paper_hierarchical:background-ood-detection}). To alleviate this, we phrase OOD detection as a likelihood ratio test of being \emph{semantically} in-distribution.
A standard likelihood ratio test \cite{buse_likelihood_1982} suggests to consider the ratio between the associated likelihoods, which we can approximate on a log-scale by the corresponding lower bounds $\mathcal{L}$ and $\mathcal{L}^{>k}$,
\begin{equation}\label{eq_hierarchical:llr-as-difference-in-likelihoods}
    % LLR = - 2 log(L0 / L1)
    % where L0 < L1 so log(L0 / L1) < 0 and LLR > 0
    % LLR = 2 log(L1 / L0)
    %     = 2 (log(L1) - log(L0))
    %     = 2 (ELBO - L^{>k})  % LLR^{>a,>b} where a=0
    % L1 = ELBO
    % L0 = L^{>k}
    LLR^{>k}(\xb) = \mathcal{L}(\xb) - \mathcal{L}^{>k}(\xb) \ .
\end{equation}
%This likelihood ratio contrasts the ELBO with $\mathcal{L}^{>k}$ for a variable choice of $k$.
Since, empirically, $\mathcal{L}\geq\mathcal{L}^{>k}$, the ratio is always positive as is standard for likelihood ratio tests.
A low value of $LLR^{>k}(\xb)$ means that the ELBO and $\mathcal{L}^{>k}$ are almost equally tight for the data.
On the contrary, a high value indicates that $\mathcal{L}^{>k}$ is looser on the data than the ELBO; hence, the data may be OOD.

We can gather further insights about this score if we write the regular ELBO and the $\mathcal{L}^{>k}$ bounds in the exact form that includes the intractable KL-divergence between the approximate and true posteriors,
\begin{align}
    \mathcal{L}      &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) || p_\thetab(\zb|\xb)\right), \label{eq_hierarchical:likelihoods-as-exact} \\ 
    \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k }|\zb_{>k}) q_\phib(\zb_{>k}|\xb) || p_\thetab(\zb|\xb)\right) \nonumber \ .
\end{align}
Subtracting these cancel out the two data likelihood terms $\log p_\thetab(\xb)$ and only the KL-divergences from the approximate to the true posterior remain,
\begin{align}
    LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) || p_\thetab(\zb|\xb)\right) \\
                 &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb) || p_\thetab(\zb|\xb)\right) \ . \notag
\end{align}\label{eq_hierarchical:llr-as-kls}
